Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-11-04 cycle/response-stumpf.tex

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\textcolor{journalname}{\textit{#1}}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}

To A.~Stumpf,

Thank you very much for your constructive review.
We are very thankful to receive feedback from the glacial geology community,
more particularly so here on The Cryosphere Discussion which traditionally
has a stronger presence in quantitative glaciology.

\referee{%
    Seguinot et al. present a well designed numerical model for the Cordilleran
    Ice Sheet (CIS) in North America for the last two glaciations occurring
    over the past 120000 years. Although the constraints on such a model are
    not yet all fully understood, their proposed simulations attempt to take
    account all the complexity of the glacial system, and utilizes a variety of
    data sets. The modeling confirms what geologists have observed in the
    field; there was a fast decay of the CIS during MIS~2 and MIS~4 and
    non-glacial conditions existed during MIS (Olympia Nonglacial Interval;
    e.g., Plouffe and Jette, 1997). I applaud them for undertaking of
    such a difficult task, and notifying the reader where input data is sparse
    or inconclusive.}

@@

Thank you very much for this positive summary of our work and for your
appreciative words!

\referee{%
    I provide the following general comments and observations which may help
    the authors in revising the manuscript for final publication. Many of these
    points are both my personal suggestions and also the recommendations of
    others currently researching the CIS or who have completed studies in the
    past.}

\referee{%
    1) With the large amount of research that has been undertaken to map the
    landforms and deposits of the CIS by federal and provincial scientists,
    academic faculty, and undergraduate and graduate students and determine the
    extent, volume and dynamics ice sheet, the impression left on the reader by
    the opening sentence would be incorrect. These studies have greatly
    advanced our understanding of the CIS, and could be an important dataset to
    test against the modeling. Stumpf et~al. (2014) provides a list of some
    of these studies.}

@@

\msquote{%
    After more than a century of geological research, the Cordilleran ice sheet
    of North America remains among the least understood in terms of its former
    extent, volume, and dynamics. Because of the mountainous topography on
    which the ice sheet formed, geological studies have often had only local or
    regional relevance, and shown such a complexity that ice sheet-wide spatial
    reconstructions of advance and retreat patterns are lacking.}

\referee{%
    2) To help the reader better understand the maps presented, I would
    recommend some spatial information be added (e.g., latitude/longitude
    grids; political boundaries; lakes and rivers, place names etc..).}

Graticules, rivers and lakes were added to all maps, the latter two as a
background to model result so that they do not interfere with them. Because we
feel that more geographic information would disturb the visibility of the main
figure contents, we have also reworked and enlarged Fig.~1 (location map) to
replace abbreviations by full geographic names, which should also help the
reader to follow discussions in the text.

\referee{%
    3) Was the model tested against regional-scale ground-based data
    (e.g., Ferbey et~al., 2013) to constrain ice divide positions, ice
    flow direction, and ice sheet thickness?}

@@

The model was tested qualitatively against an extensive body of glacial
geology literature, which is the topic of Sect.~4 (Comparison to the geologic
record). The model was not tested quantitatively against geologic evidence,
because we did not know of any publicly available dataset covering a
significant part of the model domain until reading your comment.

Therefore, we thank you very much for pointing to us the recent map and dataset
by \citet{Ferbey.etal.2013}. Indeed, it would make much sense to compare this
database against modelled basal velocities in the future. However, we feel that
such a quantitative comparison is premature at this stage. In fact, modelled
basal velocities are sensitive to hydrological properties of the subglacial
till (cf. our response to A.~Jarosch in this discussion), and even more to
thermal conditions at the base, which in turn are largely influenced by
geothermal heat flow, kept constant in this study.

\referee{%
    4) For the central sector of the CIS, Stumpf et~al. (2000) provides
    some insight into the chronology and effectiveness of glacial erosional
    during the MIS. In figure~5, and in the accompanying text, they describe
    how landforms on the surface formed. For example, in lake valleys east of
    the Skeena Mountains, it appeared the major glacial streamlined landforms
    were formed during a longer glacial advance phase, with ice flow
    paralleling the valleys, and later flows, some perpendicular the valley
    flow, only weakly impacted them.}

@@

Thank your for pointing this out. We have added a reference to
\citet{Stumpf.etal.2000} in the discussion of the erosional imprint on the
landscape (Sect.~4.2.3).

\referee{%
    5) Stumpf et~al. (2000) was the first study to extensively document a
    predominant westerly directed ice flow across high elevations in the Skeena
    and Coast Mountains. This flow direction appeared to continue into the
    late-glacial period. Other subsequent studies also confirm that
    late-glacial readvance eastward out of these mountains and retreat of ice
    margins westward into these mountains was limited.}

### 2016-11-04 cycle/abstract.tex
Move Cordillera cycle paper into subdirectory.

% Affiliations
\textsuperscript{1}{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Zürich,
          Zürich, Switzerland}\\
\textsuperscript{2}{Department of Physical Geography and the Bolin Centre for Climate
          Research, Stockholm University, Stockholm, Sweden}\\
\textsuperscript{3}{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
          Potsdam, Germany}\\

@@

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving climate
  forcing, all simulations produce two major glaciations during
  marine oxygen isotope stages~4 (61.9--56.5\,ka) and~2
  (23.2--16.8\,ka). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional oceanic sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena Mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the Last Glacial Maximum
  and hosts the last remains of Cordilleran ice until the middle Holocene
  (6.6--6.2\,ka).

>>>>

% Affiliations
\textsuperscript{1}{Laboratory of Hydraulics, Hydrology and Glaciology,
                    ETH Z\"{u}rich, Z\"{u}rich, Switzerland}\\
\textsuperscript{2}{Department of Physical Geography and the
                    Bolin Centre for Climate Research,
                    Stockholm University, Stockholm, Sweden}\\
\textsuperscript{3}{Helmholtz Centre Potsdam,
                    GFZ German Research Centre for Geosciences,
                    Potsdam, Germany}\\

@@

After more than a century of geological research, the Cordilleran
ice sheet of North America remains among the least understood in
terms of its former extent, volume, and dynamics. Because of the
mountainous topography on which the ice sheet formed, geological
studies have often had only local or regional relevance, and shown
such a complexity that ice sheet-wide spatial reconstructions of
advance and retreat patterns are
lacking. Here we use a numerical ice sheet model calibrated against
field-based evidence to attempt a quantitative reconstruction of the
Cordilleran ice sheet history through the last glacial
cycle. A series of simulations is driven by time-dependent
temperature offsets from six proxy records located around the
globe. Although this approach reveals large variations in model
response to evolving climate forcing, all simulations produce two
major glaciations during marine oxygen isotope stages 4
(62.2--56.9\,ka) and 2 (23.2--16.9\,ka). The timing of
glaciation is better reproduced using temperature reconstructions
from Greenland and Antarctic ice cores than from regional oceanic
sediment cores. During most of the last glacial cycle, the modelled
ice cover is discontinuous and restricted to high mountain
areas. However, widespread precipitation over the Skeena Mountains
favours the persistence of a central ice dome throughout the glacial
cycle. It acts as a nucleation centre before the Last Glacial
Maximum and hosts the last remains of Cordilleran ice until the
middle Holocene (6.7\,ka).
